---
title: "RI-CLPM analysis transcript: Big Five, substance use, and SES"
always_allow_html: true
output:
  github_document:
    toc: true
    html_preview: false
---

```{r setup, include=FALSE}
# echo each step's code, print messages and warnings, and keep going on errors so
# the transcript stays complete. send any figures where github renders them.
knitr::opts_chunk$set(
  echo = TRUE, message = TRUE, warning = TRUE, error = TRUE,
  comment = "#>", fig.path = "run_all_files/figure-gfm/"
)
library(magrittr) # %>%, needed by the sourced module functions

# load the pipeline through run.R, the single entry point: it sources the R/
# modules in numeric order and defines `config`. run.R's startup message is
# suppressed because this chunk is include=FALSE.
source("run.R")

# weasel drives the subset-selection showcase (section 2). attach it when the
# package is installed; otherwise that step is skipped and a fixed window is used.
use_weasel <- requireNamespace("weasel", quietly = TRUE)
if (use_weasel) suppressMessages(library("weasel", character.only = TRUE))

waves <- 5L # default window length; weasel overrides it when it runs
```


## 1 Data

```{r data}
# read the frozen per-module merges (data_merged/{cp,ch,ci}_merged.sav) plus the
# background variables (data_init/b.sav). no credentials or 2fa: the merges are
# read straight from disk.
liss <- load_liss_data(config)

# structural check: rows x cols of each module, a regression guard after any
# change to the data layer.
sapply(liss[c("p", "h", "i", "b")], function(d) {
  if (is.null(d)) c(rows = NA, cols = NA) else c(rows = nrow(d), cols = ncol(d))
})
```

```{r provenance}
# data provenance: size and last-modified time of the inputs this transcript
# reads, oldest first. modification time is portable across volumes.
inputs <- c(
  file.path(config$merged_dir, paste0(config$liss_modules, "_merged.sav")),
  config$avars_file
)
info <- file.info(inputs)
prov <- data.frame(
  file     = basename(inputs),
  contains = c("personality", "health", "income", "background"),
  size_mb  = round(info$size / 1024^2, 1),
  modified = format(info$mtime, "%Y-%m-%d"),
  age_days = round(as.numeric(difftime(Sys.time(), info$mtime, units = "days")))
)
prov[order(info$mtime), ]
```

## 2 Data Preparation

```{r clean-income}
# clean household income, then summarise what was flagged or capped.
liss$i_clean <- clean_income(liss$i, liss$b, verbose = FALSE)
print_outlier_summary(summarize_outliers(liss$i_clean))
```

```{r ses}
# construct SES as a binary median split (~50/50), the canonical coding for the
# multi-group moderation test in section 5.
liss$i_clean <- construct_ses(liss$i_clean, method = "binary")
table(liss$i_clean$ses, useNA = "ifany")
```

```{r measures}
# quick look at the constructed measures before assembly: Big Five scores from
# the personality module and substance use as a count (0 to 6).
head(compute_b5_scores(liss$p))
table(process_substance_use(liss$h, method = "sum")$substance_use, useNA = "ifany")
```

```{r assemble}
# assemble the long analysis frame, pairing personality with substance use per
# wave and carrying SES through. "sum" gives the count coding used throughout.
analysis_data <- prepare_analysis_data(liss, substance_method = "sum")
str(analysis_data, max.level = 1, list.len = 12)
```

```{r weasel, eval=use_weasel}
# principled, reproducible subset selection: restrict to waves where extraversion
# is observed, re-index onto a dense per-person axis, score scenarios, and emit a
# paste-ready methods paragraph. the chosen window length then drives the reshape.
sel <- select_weasel_subset(
  analysis_data,
  measured = "extr",
  span     = "core",
  core_len = waves
)
analysis_data <- sel$data
waves <- sel$L
cat(sel$justification, "\n")
```

```{r weasel-skip, eval=!use_weasel, echo=FALSE, results='asis'}
cat(sprintf(
  "> weasel is not installed, so subset selection is skipped; the analysis proceeds on the full assembled data with a fixed %d-wave window.\n",
  waves
))
```

```{r reshape}
# reshape long to wide for the RI-CLPM. weasel's window length sets the wave count.
data_wide <- reshape_for_riclpm(analysis_data, max_waves = waves)
cat("persons:", nrow(data_wide), "| waves:", waves, "\n")
table(data_wide$ses, useNA = "ifany")
```

## 3 Diagnostics

```{r variance}
# substance-use variance per wave. near-zero variance breaks multi-group fits;
# the count coding keeps variance well away from zero.
sub_cols <- grep("^substance_use_", names(data_wide), value = TRUE)
sapply(sub_cols, function(v) stats::var(data_wide[[v]], na.rm = TRUE))
```

```{r coverage}
# pairwise coverage of the analysis variables under FIML.
coverage <- check_coverage(
  data_wide,
  vars      = c("extr", "open", "cons", "substance_use"),
  threshold = 0.10
)
print_coverage_summary(coverage)
```

```{r group-variance}
# variance of substance use within each SES group; zero variance in a group would
# make the multi-group model fail.
check_variance_by_group(data_wide, group_var = "ses", vars = "substance_use")
```

## 4 Measurement Model

The RI-CLPM fixes every factor loading to one, so metric (equal-loading) invariance holds by
construction across waves and across SES groups. That is enough to compare the structural
parameters: the autoregressive and cross-lagged paths and the between-person random-intercept
variances. The confirmatory factor analyses below are a separate check on whether each Big Five
dimension is measured adequately by its items. Weak CFA fit signals possible attenuation of the
cross-lagged effects (estimates biased toward zero), not a violation of the RI-CLPM's loading
constraints.

```{r cfa}
cfa_results <- purrr::map(c("extr", "open", "cons"), ~ fit_b5_cfa(liss$p, .x))
cfa_fit_table <- extract_cfa_fit(cfa_results)
knitr::kable(cfa_fit_table, digits = 3, caption = "Big Five CFA fit by dimension")
```

Beyond the by-construction loading equality of the structural model, the
item-level hierarchy below tests longitudinal measurement invariance directly
on the selected occasions: configural, metric, scalar, and strict models per
trait, with fixed-factor identification and same-item residual covariances
(module 05b). The delta CFI column is the key comparison for the manuscript's
comparability claim.

```{r invariance, eval=use_weasel}
# item-level longitudinal invariance on the weasel-selected occasions; prints a
# fit table (with delta cfi) and scaled lrts per trait. `sel` is the weasel
# selection object from section 2 and carries the wave_map this needs.
invariance <- run_longitudinal_invariance(
  liss$p, sel,
  dims = c("extr", "open", "cons"), estimator = "MLR"
)
```

## 5 Hypothesis Tests

H1: personality predicts later substance use.
H2: substance use predicts later personality.
H3: socioeconomic status moderates these cross-lagged paths.

```{r hypotheses}
hypothesis_results <- run_hypothesis_tests(data_wide, waves = waves, estimator = "MLR")
print_hypothesis_summary(hypothesis_results)
```

## 6 Results and Interpretation

```{r results}
# assemble the results object the reporting functions expect, then build the
# publication tables and write them to tables/.
results <- list(
  hypothesis_results = hypothesis_results,
  cfa_fit = cfa_fit_table,
  settings = list(
    waves = waves, ses_method = "binary", substance_method = "sum",
    estimator = "MLR", use_weasel = use_weasel
  )
)
pub_table <- create_publication_table(results)

dir.create("tables", showWarnings = FALSE)
utils::write.csv(pub_table$h1_h2, "tables/h1_h2_cross_lagged.csv", row.names = FALSE)
utils::write.csv(pub_table$h3, "tables/h3_ses_moderation.csv", row.names = FALSE)
utils::write.csv(cfa_fit_table, "tables/cfa_fit.csv", row.names = FALSE)
utils::write.csv(hypothesis_results$summary, "tables/cross_lagged_summary.csv", row.names = FALSE)

knitr::kable(
  pub_table$h1_h2,
  caption = "H1/H2 within-person cross-lagged effects and random-intercept shares"
)
knitr::kable(
  pub_table$h3,
  caption = "H3 SES moderation: targeted cross-lagged multi-group test"
)
list.files("tables")
```

```{r interpret}
# narrative interpretation, including how to read null cross-lagged effects in
# the presence of strong random-intercept (between-person) stability.
interpret_riclpm_results(hypothesis_results, cfa_fit_table, coverage)
```

## Appendix: sensitivity analysis

Alternative specifications (3-level SES, binary substance use).

```{r sensitivity}
(sensitivity <- run_sensitivity_analysis(config, waves = 5))
```
